---
title: "DATASCI 306, Spring 2026, Homework 7"
author: "Your name, Your uniqname"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DBI)
library(duckdb)
library(dbplyr)
```

This homework covers material from Lectures 23--25 (**Shiny** and reactive
programming) and Lecture 28 (**databases and SQL**). It is the capstone for
the course: you will use **DuckDB** to query a data file with SQL, answer
questions about it by **joining tables**, and then wire those same queries
into a live **Shiny** app --- an interactive version of the "Six Degrees of
Kevin Bacon" game.

**Grading.** Same scheme as previous homeworks:

- **Code questions (Q1--Q3): completion.** Attempt every part in good faith.
  Code that runs and shows you engaged with the question earns full credit.
- **Reflection: points.** The reflection at the end is graded on the
  substance of your answers (10 points total).

Question 3 is built in a separate `.R` file (a Shiny app cannot be knit into
a PDF), which is pulled into this document automatically. The credit there is
in completing the app, which you test by running it.

## Background: the data

The file `data/movie_actors.parquet` is a table of actor--movie pairs: each row
records that a particular `actor` appeared in a particular `movie`. Movie
titles carry their release year as a suffix, e.g. `Footloose (1984)`.

Recall from Lecture 28 that **DuckDB can query a file directly**, and that
**Parquet** is a columnar, compressed, binary file format. You point `FROM` at
the `.parquet` file and DuckDB reads only what it needs, without importing
anything into the database. We will do everything that way --- query the file
in place rather than loading it into R: just `FROM 'data/movie_actors.parquet'`.

---

## Question 1 (3 points): Querying a file with DuckDB

**(a)** Open an in-memory DuckDB connection (this is just the query *engine* ---
no data is stored in it). Then take a first look at the file by selecting its
first 5 rows directly.

```{r}

```

**(b)** Using **SQL**, answer these warm-up questions about the file:

1. How many distinct actors and how many distinct movies are in the data?
2. How many movies does Kevin Bacon appear in?
3. Which five movies have the largest casts?

```{r}

```

**(c)** List the movies Kevin Bacon appeared in, **oldest first**. The year is
the four digits inside the `(YYYY)` suffix; pull it out in SQL with
`regexp_extract(movie, '(\d{4})\)$', 1)` and `TRY_CAST(... AS INTEGER)`.

```{r}

```

---

## Question 2 (5 points): Joining tables

This is the heart of the homework. Two actors are **co-stars** if they appear
in the *same* movie --- which we find by joining the table **to itself**.

**(a) A self-join.** Join `'data/movie_actors.parquet'` to itself: alias one copy
`a` and another `b`, match rows with the same `movie` but a different `actor`,
and the `b.actor` values are the co-stars of the `a.actor` values. Find Kevin
Bacon's top 10 co-stars --- the actors who share the most movies with him.

```{r}

```

**(b) The same join, no `WHERE`.** Drop the filter on Kevin Bacon and ask the
question for *everyone*: which 10 actors have the most distinct co-stars in the
whole data set? (Use `COUNT(DISTINCT b.actor)`.)

```{r}

```

**(c) The same query with `dbplyr`.** Writing SQL by hand is one option;
`dbplyr` is the other. Give the file a name DuckDB can refer to with a **view**
(a saved query --- it still reads the file, nothing is copied in), point
`tbl()` at it, and write the co-star query as a `dplyr` pipeline. Confirm the
generated SQL with `show_query()`, then `collect()` the top 10.

```{r}

```

---

## Question 3 (6 points): The Six Degrees of Kevin Bacon game

The **Six Degrees of Kevin Bacon** is a trivia game built on the claim that any
film actor can be linked to Kevin Bacon through a short chain of shared movie
appearances: an actor's "Bacon number" is the number of steps in the shortest
such chain (Bacon himself is 0, anyone who acted with him is 1, and so on). The
game has a well-known online version, the **Oracle of Bacon**
(<https://oracleofbacon.org/>), and its own
[Wikipedia article](https://en.wikipedia.org/wiki/Six_Degrees_of_Kevin_Bacon).

You will build a playable version as a **Shiny** app. The player starts from a
random actor and walks toward Kevin Bacon: pick an actor, see their movies;
pick a movie, see its cast; repeat until you reach Bacon (win) or run out of
six steps (lose).

A starter file `bacon_app.R` is provided. Its data section is already written
for you, and it is worth reading closely: the helper functions `get_movies()`
and `get_cast()` are **SQL queries that DuckDB runs directly against the
Parquet file** --- the same `FROM 'data/movie_actors.parquet'` pattern from
Questions 1--2.
This is the payoff of the hybrid: the app's data layer is SQL, and the
interactivity is Shiny.

**Your job is to fill in the server logic** --- the four `# TODO` blocks inside
the `server` function. Do not modify the data section, the helpers, or the UI.

1. **TODO 1 --- `observeEvent(input$new_game, ...)`:** Reset the game state and
   sample 5 random starting actors, weighted by number of movies (pass
   `prob = actor_weights` to `sample()`). Use `updateRadioButtons()` to set the
   choices for `"selection"`.

2. **TODO 2 --- `observeEvent(input$choose, ...)`:** The main game logic. In the
   `"actor"` phase, record the actor, increment the step, check win/loss, then
   look up their movies with `get_movies()` (minus `game$visited_movies`) and
   switch to the `"movie"` phase. In the `"movie"` phase, record the movie (and
   add it to `game$visited_movies`), look up the cast with `get_cast()` (minus
   `game$visited`), and switch back to `"actor"`.

3. **TODO 3 --- `output$path_display`:** With `renderUI()`, show the path so far,
   actors in **bold** and movies in *italics*, separated by arrows.

4. **TODO 4 --- `output$result_message`:** With `renderText()`, show a win
   message (with the Bacon number), a loss message, or nothing.

### Provided helpers (already in the file)

- `get_movies(actor_name)` --- a SQL query returning an actor's movies, oldest
  first
- `get_cast(movie_name)` --- a SQL query returning the actors in a movie
- `non_bacon_actors` --- every actor except Kevin Bacon
- `actor_weights` --- movie counts per actor (use as `prob =` when sampling)

### Hints

- `updateRadioButtons(session, "selection", choices = new_vector)` changes the
  options; `session` is a parameter of the server function.
- `game$phase` is `"actor"` or `"movie"` --- check it to decide what to do.
- `game$visited` / `game$visited_movies` track what has been used; remove them
  with `setdiff()`.
- `game$step` counts *actor* selections only; the game ends at 6.
- Guard the choose handler with `req(game$active, input$selection)`.
- For the path, odd indices of `game$path` are actors, even indices are movies.

Test with `shiny::runApp("bacon_app.R")`. Your completed code is included
automatically below:

```{r, code=readLines("bacon_app.R"), eval=FALSE}
```

---

## Reflection (10 points)

This section is graded on the substance of your answers, not on getting the
"right" answer. Aim for a few honest sentences each. One-word answers will not
earn full credit.

### 1. Hardest moment (3 points)

Which part of this homework took the longest, or where did you get most stuck?
Describe specifically what was confusing and how you got unstuck.

### 2. SQL vs. doing it in R (3 points)

In Question 2 you found co-stars with a SQL self-join that DuckDB runs against
the file. You could instead read the whole file into R first and compute
co-stars with `dplyr`. For a file of a few hundred million rows, which approach
is better, and why? Use the words *lazy* and *collect* in your answer.

### 3. Reactivity (2 points)

The Shiny app stores the game in `reactiveValues` and changes the radio-button
choices from inside an `observeEvent`. In your own words, why does the app need
*reactive* state here instead of ordinary R variables?

### 4. A question for class (2 points)

One question you would want to ask in class about SQL, databases, or Shiny.

---

### What to submit

- This file, knit to PDF (`hw7_sp26.Rmd` --> `hw7_sp26.pdf`).
- Your completed `bacon_app.R`.
