---
title: "DATASCI 306, Spring 2026, Homework 5"
author: "Your name, Your uniqname"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

This homework covers material from Lectures 17, 18, and 19: applying
**regular expressions** to select, detect, and extract text
(`matches`, `str_detect`, `str_extract_all`, alternation, backreferences);
**factors** with the `forcats` package
(`factor`, `levels`, `fct_infreq`, `fct_reorder`, `fct_collapse`,
`fct_lump_n`); and **iteration** with `for`/`while` loops and the
`purrr` `map` family (`map`, `map_*`, `map2_*`, the `~ .x` formula
shorthand, `safely`, and `transpose`).

**Grading.** Same scheme as previous homeworks:

- **Code questions (Q1--Q3): completion.** Attempt every part in good
  faith. Code that runs and shows you engaged with the question earns
  full credit.
- **Reflection: points.** The reflection at the end is graded on the
  substance of your answers (10 points total).

### Submission checklist

Before submitting, confirm:

- [ ] You replaced the name and uniqname in the YAML header.
- [ ] The document knits to PDF without errors.
- [ ] Every R chunk has been run and shows output where output is expected.
- [ ] You answered the reflection at the bottom of the document.

## Question 1: Regular expressions on congressional tweets

The U.S. Congress is seated in two-year periods; the 115th Congress sat
in 2017--2018. Researchers at UM's School of Information collected the
tweets that sitting members of Congress (MoCs) posted during this period
and flagged which were **retweets** of other accounts. We will work with
a sample of that retweet data. (The full text of each retweet is in the
`full_text` column.)

```{r}
retweets <- read_csv(
  "data/retweets_congress_115th_sample.csv.gz",
  col_types = cols(created_at = col_datetime()),
  show_col_types = FALSE
)
dim(retweets)
retweets |> select(user_screen_name, chamber, party, full_text) |> head(3)
```

### Part a --- Selecting columns with a name regex

In Lecture 17 we used `matches()` to pick columns whose **name** fits a
regular expression. Three columns describe the retweet relationship --- the
original account and its reach: `retweet_count`, `rt_screen_name`, and
`rt_followers_count`. Using `select(retweets, matches(...))` with a single
regex, select **exactly** these three columns and report their names.


```{r}

```

### Part b --- Detecting a pattern with `str_detect`

The [American Health Care Act](https://en.wikipedia.org/wiki/American_Health_Care_Act_of_2017)
(AHCA) was the 2017 bill to partly repeal the Affordable Care Act (ACA).
Tweets about it used the hashtags `#AHCA` and `#ACA`.

Write a **single** regular expression and use `str_detect` to add a
logical column `health_bill` that is `TRUE` when `full_text` contains
either `#AHCA` or `#ACA`. How many tweets match?

Then group the matching tweets by month and report which month had the
most. *Hint: `floor_date(created_at, "month")` collapses a timestamp to
the first of its month; alternation `#(AHCA|ACA)` matches either hashtag.*

```{r}

```

### Part c --- Extracting all matches with `str_extract_all`

A hashtag is a `#` followed by one or more letters (`#[A-Za-z]+`). Use
`str_extract_all` to pull **every** hashtag out of `full_text`, `unlist`
the result into one character vector, lower-case it, and report the 10
most common hashtags. (`str_extract` would only give the first hashtag in
each tweet --- you want all of them.)

```{r}

```

### Part d --- Building a pattern with a backreference

We will find tweets with an accidental doubled word --- a stray "the the"
or an emphatic "very very". Rather than give you the finished pattern,
here are the pieces to assemble.

Remember that in an R string every regex backslash must be **doubled**: you
write `"\\w"` in your code for the regex `\w`. The pieces:

- `\\w+` --- one or more **word characters** (letters, digits, or `_`).
- `(...)` --- a **capturing group**: parentheses remember the text they
  matched so you can refer to it later. Wrapping the word as `(\\w+)`
  captures it as group 1.
- `\\1` --- a **backreference**: it refers back to whatever group 1
  captured, i.e. *the same word again*. This is the piece that makes the
  pattern look for a repeat rather than just any two words.
- `\\s+` --- one or more **whitespace** characters (the gap between the
  two words).
- `\\b` --- a **word boundary** (the edge between a word and a non-word
  character), so you match whole words and not fragments of longer ones.

Combine these into a single pattern that matches a word, then whitespace,
then that **same** word repeated, with a word boundary on each end. Then
use `str_detect` to count how many tweets contain an immediately repeated
word, and print a few with `str_subset`.

*Hints: the order of the pieces is boundary, captured word, whitespace,
backreference, boundary. Wrap your finished pattern in*
`regex(your_pattern, ignore_case = TRUE)` *so that "The the" also counts.*

```{r}

```

## Question 2: Factors and the WikiArt-Emotions data

The WikiArt-Emotions data set (the one from lecture) records, for each
work of art, average annotator ratings of how strongly the piece evokes
each of 20 emotions, under three viewing conditions. The chunk below
loads it and reshapes it into the tidy long form you saw in lecture ---
you do **not** need to change it.

```{r}
wikiart <- read_tsv("data/WikiArt-Emotions-All.tsv.gz", show_col_types = FALSE) |>
  mutate(Year = suppressWarnings(as.numeric(str_sub(Year, 1, 4))))

wa_long <- wikiart |>
  pivot_longer(
    matches(":"),
    names_to  = c("rating_type", "emotion"),
    names_sep = ":",
    values_to = "mean_annotation"
  ) |>
  mutate(emotion = str_replace(emotion, "\\s", ""))

select(wa_long, ID, Title, rating_type, emotion, mean_annotation)
```

### Part a --- The strongest emotion per piece

For each work of art, we want the single emotion rated most strongly
under the `"Art (image+title)"` viewing condition. Starting from
`wa_long`:

- keep only rows whose `rating_type` mentions `"Art"` (use `str_detect`),
- group by `ID`, and for each piece keep the `emotion` with the largest
  `mean_annotation` (the `arrange(desc(...))` + `first()` trick from
  lecture). Call the result `wa_strong`, with columns `ID`, `strongest`,
  and `strongest_val`.

Then make a bar chart of how often each emotion is the strongest, with
the bars ordered from most to least common. *Hint: `fct_infreq` orders a
factor by frequency.*

```{r}

```

### Part b --- Lumping a long tail with `fct_lump_n`

The bar chart in part a has a long tail: many emotions are the strongest
for only a handful of pieces. The `forcats` package has a family of
`fct_lump_*` functions for exactly this situation --- folding rare levels
together into a single catch-all. We did not see these in lecture, so take
a look at the help page with `?fct_lump_n` first.

Use `fct_lump_n` to keep the **6** most common strongest-emotions and
collapse every other emotion into a single `"Other"` level. Add this as a
new column `emo_lumped` on `wa_strong` and report the count of pieces per
level. Roughly what fraction of pieces land in `"Other"`?

```{r}

```

### Part c --- Reordering by a summary with `fct_reorder`

Do more highly rated works tend toward particular emotions? Join the
overall `Ave. art rating` column from `wikiart` onto `wa_strong` (by
`ID`), then for each `strongest` emotion compute the mean art rating.

Make a dot plot (`geom_point`) of mean art rating by emotion, with the
emotions **ordered by their mean rating** using `fct_reorder`. Which
emotion has the highest mean art rating, and which the lowest?

```{r}

```

### Part d --- Extracting and lumping a messy category

The `Category` column packs one or more art styles into a single
comma-separated string (e.g. `"Abstract Art,Cubism"`). Use a regular
expression with `str_extract` to pull out just the **primary** (first)
style --- everything up to the first comma --- into a new column
`primary_style`. *Hint:* `"^[^,]+"` matches the run of non-comma
characters at the start of the string.

There are many distinct styles, but this collection is fairly balanced
(roughly 200 works per style). So instead of keeping a fixed number of
levels, use `fct_lump_min` to lump together any style appearing fewer
than **150** times into `"Other"`, and report the counts per lumped
style. Which styles were rare enough to be lumped?

```{r}

```

## Question 3: Iteration and the `map` family

### Part a --- A `for` loop over columns

The `ggplot2::mpg` table (loaded with the tidyverse) has columns of
mixed type. Using a `for` loop --- **not** `map` --- build a named
character vector `col_classes` giving the `class` of each column of
`mpg`. Pre-allocate the storage with `vector("character", ncol(mpg))`
and loop with `seq_along`. (Recall a table is a list of columns, so
`mpg[[i]]` is the i-th column.)

```{r}

```

### Part b --- The same thing with `map`, plus a custom summary

1. Reproduce part a in a single line using `map_chr`.
2. Write a function `col_summary(x)` that returns a short string: for a
   **numeric** column, the rounded mean (e.g. `"num: mean=3.5"`); for any
   other column, the number of distinct values (e.g. `"cat: 7 levels"`).
   Use `map_chr` to apply it to every column of `mpg`.

```{r}

```

### Part c --- Mapping over a list of samples

A **compound random variable** draws a random sample size and then draws
that many values. The function below returns a list of `n` numeric
vectors of varying lengths:

```{r}
r_pois_norm <- function(n, lambda = 1, mu = 0, sd = 1) {
  replicate(n, {
    n_i <- rpois(1, lambda) + 1
    rnorm(n_i, mu, sd)
  })
}

set.seed(306)
samples <- r_pois_norm(6, lambda = 8)
```

Using the `map` family (and the `~ .x` formula shorthand where handy):

1. report the length of each sample as an integer vector (`map_int`);
2. report the largest **absolute** value within each sample as a numeric
   vector (`map_dbl`);
3. for a second list of samples `samples_y` (built below), compute the
   mean ratio `.x / .y` within each pair of samples using `map2_dbl`.

```{r}
samples_y <- map(samples, ~ runif(length(.x)))
```

```{r}

```

### Part d --- Errors in iteration with `safely`

Iterating over messy data, one bad element can blow up the whole loop:

```{r}
x <- list(1, 10, "3", 7)
```

```{r eval=FALSE}
map(x, log)   # errors on the character element "3"
```

Use `safely(log)` with `map` to apply `log` to every element of `x`
without crashing. Then use `transpose` to split the results into a
`$result` part and an `$error` part, and report which element(s) of `x`
caused an error.

```{r}

```

## Reflection (10 points)

This section is graded on the substance of your answers, not on getting
the "right" answer. Aim for a few honest sentences each. One-word
answers will not earn full credit.

### 1. Hardest moment (3 points)

Which part of this homework took the longest, or where did you get most
stuck? Describe specifically what was confusing and how you got unstuck.

### 2. Reading a regex (3 points)

Pick one regular expression you wrote in Question 1 and explain, piece by
piece, what each part of the pattern matches. Pretend you're explaining
it to a classmate who has never seen a backreference.

### 3. Loop vs. map (2 points)

In Question 3 you wrote the same column-class computation as both a `for`
loop (a) and a `map_chr` call (b). In a sentence or two, when do you
think a `for` loop is the clearer choice, and when is `map` clearer?

### 4. A question for class (2 points)

One question you would want to ask in class about regular expressions,
factors, or iteration.
