---
title: "DATASCI 306, Spring 2026, Homework 2"
author: "Your name, Your uniqname"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
aatemp <- read_csv("data/ann_arbor_weather.csv.gz", show_col_types = FALSE)
spotify <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-01-21/spotify_songs.csv",
  show_col_types = FALSE
) |>
  mutate(track_album_release_year = as.numeric(substr(track_album_release_date, 1, 4))) |>
  select(-c(track_name, track_album_name, playlist_name, track_artist, track_album_release_date))
```

This homework covers material from our second week of class: the structure
of tables, the core `dplyr` verbs (`filter`, `select`, `mutate`,
`group_by`, `summarize`, `arrange`), grouped transformations, and
multi-level aggregation, plus more practice with `ggplot2`.

**The data.** Two data sets:

1. A record of daily weather observations from a station in Ann Arbor,
   Michigan (`aatemp`), used in Question 1.
2. A modified version of the
   [2020-01-21 TidyTuesday](https://github.com/rfordatascience/tidytuesday)
   Spotify songs data set, used in Questions 2--7. A description of each
   column is at the
   [TidyTuesday README](https://github.com/rfordatascience/tidytuesday/blob/main/data/2020/2020-01-21/readme.md).
   This version converts `track_album_release_date` to
   `track_album_release_year` and drops a few text-heavy columns.

```{r, results = "hide"}
colnames(spotify)
```

**Grading.** Same scheme as Homework 1:

- **Code questions (Q1--Q7): completion.** Attempt every part in good
  faith. Code that runs and shows you engaged with the question earns full
  credit.
- **Reflection: points.** The reflection at the end is graded on the
  substance of your answers (10 points total). This is where careful
  thinking pays off.

### Submission checklist

Before submitting, confirm:

- [ ] You replaced the name and uniqname in the YAML header.
- [ ] The document knits to PDF without errors.
- [ ] Every R chunk has been run and shows output where output is expected.
- [ ] You answered the reflection at the bottom of the document.

## Question 1: Getting to know a table (Ann Arbor weather)

We'll start with the `aatemp` table. The columns we will work with most
are:

- `DATE` --- date of the observation
- `TMAX`, `TMIN` --- the day's maximum and minimum temperatures (deg. F)
- `PRCP` --- precipitation (inches)
- `SNOW`, `SNWD` --- snowfall and snow depth (inches)

### Part a: Rows, columns, names

Report the number of rows and the number of columns in `aatemp`. Then list
all column names. (Hint: `nrow`, `ncol`, `colnames`.)

```{r}

```

### Part b: First and last observations

Show the first 5 rows and the last 5 rows of `aatemp`. What is the earliest
date in the data? What is the latest?

```{r}

```

### Part c: Built-in summary

Run `summary` on `aatemp`. Use the output to answer:

- What was the highest `TMAX` ever observed in this record?
- What was the lowest `TMIN`?
- How many missing values are there in the `SNOW` column?

```{r}

```

### Part d: Targeted summaries

Without using `summary`, compute the mean, median, and standard deviation
of `TMAX`. Then compute the *proportion* of days in the record that had any
measurable precipitation (`PRCP > 0`). (Hint: the mean of a logical vector
is the proportion of `TRUE` values; remember `na.rm = TRUE`.)

```{r}

```

## Question 2: Filtering to a recent date range (Spotify)

Filter the Spotify data set to include only songs with
`track_album_release_year` between 1990 and 2019 (inclusive). Save the
filtered results into a new table called `spotify2`. This filtered table
will be used for the remainder of the assignment.

Using `spotify2`, determine the number of songs released in each year.

```{r}

```

## Question 3: Trends in song genres over time

After completing Question 2, enable the following code chunk by setting
`eval = TRUE`:

```{r eval = FALSE}
table(spotify2$playlist_genre)
```

Group `spotify2` by `track_album_release_year` and `playlist_genre`, and
count the number of songs released for each genre in each year. Using
these results, create a line plot where each genre is represented by a
separate line, showing how the number of released songs for each genre has
changed over time.

Based on the graph, identify which genre has shown the most significant
increase over the years.

```{r}

```

## Question 4: `mutate` on the whole table

The `cut` function breaks a quantitative variable into discrete
categories. Here is an example:

```{r}
x <- c(2, 0, 5, 10, 3, 4, -0.25, 6)
cut(x, breaks = c(-Inf, 0, 5, Inf), labels = c("Small", "Medium", "Large"))
```

The notation `(a, b]` means a half-open interval defined by $a < x \le b$.

### Part a: Energy groups

Using `cut` inside a single call to `mutate`, add a new column
`energy_group` to `spotify2` that breaks `energy` into the following
groups:

* Low Energy: `(0, 0.25]`
* Medium Low Energy: `(0.25, 0.5]`
* Medium High Energy: `(0.5, 0.75]`
* High Energy: `(0.75, 1]`

```{r}

```

### Part b: Several new columns at once

In a single call to `mutate`, add the following additional columns to
`spotify2` (keep the `energy_group` column from Part a):

- `decade`: the decade of release, as a character string (`"1990s"`,
  `"2000s"`, `"2010s"`). One way: `paste0(10 * (year %/% 10), "s")`.
- `duration_min`: the song's duration in minutes (`duration_ms / 60000`).
- `loud_per_min`: `loudness` divided by `duration_min`.

Show the first few rows of `spotify2` to confirm all three new columns
appear alongside `energy_group`. Save the updated `spotify2` so the new
columns are available for later questions.

```{r}

```

### Part c: A summary using the new columns

Using `group_by` and `summarize`, compute the count of songs in each
`energy_group`. Which energy level is most common? (Hint: `n()`.)

```{r}

```

## Question 5: Aggregating at more than one level

In this question you will summarize the data at two different levels of
grouping --- first by `playlist_subgenre` and year, then collapsing the
result up to `playlist_subgenre` and `playlist_genre`.

### Part a: Year-by-subgenre counts

Group `spotify2` by `playlist_subgenre` and `track_album_release_year`,
and use `n()` to count the songs released in each subgenre in each year.
Save this result as `by_subgenre_year`.

```{r}

```

### Part b: Average yearly releases per subgenre

Starting from `by_subgenre_year`, group by `playlist_subgenre` only and
compute the average number of songs released per year for each subgenre.
Arrange in descending order and report the top three subgenres.

```{r}

```

### Part c: One level up --- to genre

For each `playlist_genre`, compute *both* the average yearly release count
across the subgenres in that genre *and* the number of distinct subgenres
in the genre. You may start from your Part b result or build a fresh
pipeline. (Hint: `n_distinct` is useful.)

Arrange the result by descending average yearly releases.

```{r}

```

## Question 6: `mutate` *within* groups

Sometimes you want to compute a per-group summary but keep the row-level
structure of the table. This is what grouped `mutate` does: unlike
`summarize`, `mutate` keeps every row, but the summary value is computed
*within* each group and broadcast to every row of the group.

### Part a: Genre-relative popularity

For each song in `spotify2`, compute its popularity *relative to the
average popularity of its genre*. Specifically, add a new column
`popularity_vs_genre` defined as

$$
\texttt{popularity\_vs\_genre} = \texttt{track\_popularity}
  - \overline{\texttt{track\_popularity}}_{\text{ genre}}
$$

using `group_by(playlist_genre)` together with `mutate`. Don't forget to
`ungroup()` afterwards.

```{r}

```

### Part b: Most over-performing song per genre

Using your new column, find the single song in each genre with the
*largest* positive `popularity_vs_genre`. Report the genre, the subgenre,
the track id, the raw `track_popularity`, and `popularity_vs_genre`. (Hint:
`slice_max`.)

```{r}

```

### Part c: A check

For each genre, what is the average of `popularity_vs_genre`? Briefly
explain (one sentence) why your answer makes sense.

```{r}

```

> *Your explanation here.*

## Question 7: What makes a genre "danceable"? (AI as collaborator)

In this question you will explore how `danceability` relates to other
audio features --- specifically `energy`, `tempo`, and `valence` ---
across different genres, *with an AI assistant as a coding partner*.

The end goal is:

1. For each genre, compute the correlation between `danceability` and each
   of `energy`, `tempo`, and `valence`.
2. Display the result as a single visualization that makes comparison
   across genres and across the three audio features easy to read.

### Part a: Your prompt

Open [UMGPT](https://umgpt.umich.edu) (or another AI assistant of your
choice). Write a prompt that gives the AI enough context to produce code
that runs against `spotify2`. Your prompt should:

- name the data set and the relevant columns,
- state both the calculation and the visualization goal clearly,
- ask for tidyverse code (`dplyr` + `ggplot2`), not a general explanation.

Paste your prompt verbatim below.

> *Your prompt here.*

### Part b: The AI's code

Paste the code the AI produced into the chunk below and run it. If it does
not run as-is, fix it, and briefly note what you had to change. (You do
not need to keep the AI's narration --- just the code.)

```{r}

```

### Part c: Your interpretation

In two or three sentences, answer the original question: what combination
of features seems to make a song danceable, and does that combination
differ across genres? Refer to specific numbers or specific features of
your plot.

> *Your answer here.*

### Part d: Critique

In one or two sentences, evaluate the AI's code. Did it pick reasonable
`dplyr` verbs? Did its visualization make the comparison easy, or would
you change the geom, color mapping, or layout? What would you do
differently if you redid this from scratch?

> *Your critique here.*

## Reflection (10 points)

This section is graded on the substance of your answers, not on getting
the "right" answer. Aim for a few honest sentences each --- enough that
someone reading it could understand what you actually experienced.
One-word answers will not earn full credit.

You may use AI on this section, but do so in a way that supports your own
thinking rather than supplants it.

### 1. Hardest moment (3 points)

Which question took the longest, or where did you get most stuck?
Describe specifically what was confusing (a verb, a concept, an error
message, a prompt that wouldn't behave) and how you got unstuck.

### 2. `summarize` vs. grouped `mutate` (3 points)

Question 5 used `summarize` to collapse the table to one row per group.
Question 6 used `mutate` to add a per-group value to *every* row. In your
own words, when would you reach for one rather than the other? Give a
concrete example (it does not have to be from this homework).

### 3. What's still fuzzy (2 points)

Name one concept from this week (multi-level aggregation, grouped
`mutate`, `cut`, line plots over time, etc.) that you can sort of use but
could not yet explain to someone else. Be specific.

### 4. A question for class (2 points)

One question you would want to ask in class about this material. It can
be about `ggplot2`, `dplyr`, your AI interaction in Q7, or how any of
this connects to data science more broadly.
