---
title: "DATASCI 306, Spring 2026, Homework 3"
author: "Your name, Your uniqname"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(Lahman)
aatemp <- read_csv("data/ann_arbor_weather.csv.gz", show_col_types = FALSE)
```

This homework covers material from Lectures 7, 8, and 12: writing and
composing functions, `across`, predicates, anonymous functions, curly-brace
`{{}}` column arguments, variable-length `...` arguments, lexical scoping,
table types, loading delimited data, and working with dates using `lubridate`.

**Grading.** Same scheme as previous homeworks:

- **Code questions (Q1--Q3): completion.** Attempt every part in good faith.
  Code that runs and shows you engaged with the question earns full credit.
- **Reflection: points.** The reflection at the end is graded on the
  substance of your answers (10 points total).

### Submission checklist

Before submitting, confirm:

- [ ] You replaced the name and uniqname in the YAML header.
- [ ] The document knits to PDF without errors.
- [ ] Every R chunk has been run and shows output where output is expected.
- [ ] You answered the reflection at the bottom of the document.

## Question 1: Detecting and visualizing outliers

The functions you write below will use a threshold `k` — the number of
standard deviations beyond which a value is considered an outlier — from
the surrounding environment rather than as a function argument.

### Part a

Compute `k` as the 99th percentile of the absolute Z-scores across all
numeric columns of `aatemp`. Use `scale` (which standardizes a matrix
column-by-column), then take absolute values and call `quantile`. Print `k`.

Then write a function `is_outlier(x)` that returns a logical vector the
same length as `x`: `TRUE` where a value is more than `k` standard
deviations from the mean of `x`, and `FALSE` otherwise. Use `na.rm = TRUE`
when computing the mean and standard deviation.

Using `aatemp`, add an `outlier` column with `mutate` and make a scatter
plot of `PRCP` over `DATE`, with the color of each point determined by
whether it is an outlier.

```{r}

```

### Part b

Write a predicate `has_outlier(x)` that returns a single `TRUE` or `FALSE`
indicating whether `x` contains at least one outlier. (Hint: use
`is_outlier` and `any`; handle `NA`s carefully.)

Use `has_outlier` with `where` inside `select` to find which numeric columns
of `aatemp` contain at least one outlier. Report the column names.

```{r}

```

### Part c

Write a function `plot_col(tbl, col)` that takes a data frame and a column
name (using `{{col}}` curly-brace syntax). Inside the function:

- Extract the column's values with `pull` and call `has_outlier` on them.
- **If the column has outliers:** produce a histogram where the fill color
  indicates whether each observation is an outlier.
- **If the column has no outliers:** produce a density plot.

Demonstrate by calling `plot_col` on `PRCP` (which has outliers) and on
`TMAX` (which does not).

```{r}

```

## Question 2: Loading data and working with dates

The CDC publishes monthly counts of cancer deaths broken down by type and
demographic group. The file
`AH_Provisional_Cancer_Death_Counts_by_Month_and_Year__2020-2021.csv.gz`
is in your `data/` folder. Each row covers one reporting period: the three
date columns `Data As Of`, `Start Date`, and `End Date` are stored as
`MM/DD/YYYY` strings.

### Part a

Read the file into a tibble called `cancer_deaths`. Use `col_types` and
`col_date("%m/%d/%Y")` to parse all three date columns directly on load.
Confirm the column types look correct by printing the first few rows.

```{r}

```

### Part b

Using the date columns you just loaded, answer the following with
`lubridate`:

1. What is the full time span of the data set, from the earliest `Start
   Date` to the latest `End Date`? Report it in months.
2. How long is each individual reporting period (`End Date - Start Date`
   in days)? Are all periods the same length, or does it vary?
3. Do any `Start Date` values fall on a day of the week other than Sunday?
   (Use `wday`.)

```{r}

```

## Reflection (10 points)

This section is graded on the substance of your answers, not on getting the
"right" answer. Aim for a few honest sentences each. One-word answers will
not earn full credit.

### 1. Hardest moment (3 points)

Which part of this homework took the longest, or where did you get most
stuck? Describe specifically what was confusing and how you got unstuck.

### 2. Building functions in layers (3 points)

In Question 1 you wrote `is_outlier`, then built `has_outlier` on top of
it, then built `plot_col` on top of that. In one or two sentences, describe
one advantage of this layered approach compared to writing a single function
that does everything at once.

### 3. Dates in data (2 points)

In Question 2, the date columns in the cancer deaths file were stored as
strings and had to be parsed with a format string. Name one thing that can
go wrong if you skip this step and just leave dates as character columns.

### 4. A question for class (2 points)

One question you would want to ask in class about function writing, date
handling, or anything else from this homework.
