---
title: "DATASCI 306, Spring 2026, Homework 4"
author: "Your name, Your uniqname"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Lahman)
```

This homework covers material from Lectures 13, 14, and the set-operations
portion of Lecture 15: tidy vs. wide data, `pivot_longer`/`pivot_wider`,
relational data, primary and foreign keys, mutating joins (left/right/full/
inner), filtering joins (`semi_join`, `anti_join`), compound keys, natural
joins, and the base R set-operation functions (`intersect`, `union`,
`setdiff`, `setequal`, `%in%`).

**Grading.** Same scheme as previous homeworks:

- **Code questions (Q1--Q3): completion.** Attempt every part in good faith.
  Code that runs and shows you engaged with the question earns full credit.
- **Reflection: points.** The reflection at the end is graded on the
  substance of your answers (10 points total).

### Submission checklist

Before submitting, confirm:

- [ ] You replaced the name and uniqname in the YAML header.
- [ ] The document knits to PDF without errors.
- [ ] Every R chunk has been run and shows output where output is expected.
- [ ] You answered the reflection at the bottom of the document.

## Question 1: Pivoting cancer death counts

You loaded the CDC's monthly cancer-death file in Homework 3. Here we will
reshape it and answer some questions about it. The same file lives in this
homework's `data/` folder; load it the same way you did before.

```{r}
cancer_deaths <- read_csv(
  "data/AH_Provisional_Cancer_Death_Counts_by_Month_and_Year__2020-2021.csv.gz",
  col_types = cols(
    "Data As Of" = col_date("%m/%d/%Y"),
    "Start Date" = col_date("%m/%d/%Y"),
    "End Date"   = col_date("%m/%d/%Y")
  ),
  show_col_types = FALSE
) |>
  select(-`Malignant neoplasms (C00-C97)`)
```

The `Malignant neoplasms (C00-C97)` column is a total across all cancer
types and has been dropped above so the remaining cancer columns sum to a
sensible total.

Because the cancer-type column names are very long, you may find it useful
to print tables with `knitr::kable`, e.g.

```{r}
select(cancer_deaths, 1:5) |> head(5) |> knitr::kable()
```

### Part a — Tidy the data with `pivot_longer`

Each row of `cancer_deaths` is a (reporting period, demographic) combination,
and each of the many cancer-type columns holds a count. That is *wide*
format: the cancer type is encoded in the column name rather than living in
a column of its own.

Use `pivot_longer` to reshape `cancer_deaths` so there is one row per
(reporting period, demographic, cancer type) with the death count in a
single `deaths` column. The non-cancer columns (dates, sex, race, age group,
etc.) should be carried along untouched. Call the result `cancer_tidy` and
give the new columns informative names (e.g. `cancer_type` and `deaths`).

Hint: see the `tidy-select` helpers — `where(is.numeric)` or a `-c(...)`
selection of the columns you do *not* want to pivot are both reasonable.

```{r}

```

### Part b — Summaries that are easier on the tidy version

Using `cancer_tidy`, answer the following. Each part should be a short
`group_by` / `summarize` / `filter` pipeline.

1. How many cancer deaths are recorded in this data set in total?
2. Group by `End Date` (the end of the reporting period) and compute the
   total deaths in each month. Which month had the most deaths?
3. Restrict to the following five cancer types and graph total monthly
   deaths over time, with one line per cancer type. (The hint
   `theme(legend.position = "bottom")` keeps the long category names from
   crowding the plot.)

```{r}
cancer_type_subset <- c(
  "Malignant neoplasms of trachea, bronchus and lung (C33-C34)",
  "All other and unspecified malignant neoplasms (C17,C23-C24,C26-C31,C37-C41,C44-C49,C51-C52,C57-C60,C62-C63,C66,C68-C69,C73-C80,C97)",
  "Malignant neoplasms of lymphoid, hematopoietic and related tissue (C81-C96)",
  "Malignant neoplasms of colon, rectum and anus (C18-C21)",
  "Malignant neoplasm of pancreas (C25)"
)
```

```{r}

```

### Part c — Sparse vs. dense

Many (period, demographic, cancer type) combinations had zero recorded
deaths. If we agreed to treat *missing* rows as zero, we could drop those
rows entirely.

- How many rows does `cancer_tidy` have? How many of those rows have
  `deaths == 0`?
- Build `cancer_tidy_sparse` by filtering out the zero-death rows. What
  percentage smaller is the sparse version than the full tidy table?
- Using `cancer_tidy_sparse`, report the top 5 cancer types by total
  deaths across the whole data set.

```{r}

```

### Part d — Faceted percentages by age group

Using `cancer_tidy_sparse` restricted to the five cancer types in
`cancer_type_subset`, build a faceted bar plot showing, for each `Age
Group`, the *percentage* of deaths from those five cancers attributable
to each individual cancer type. (In other words: within each age-group
facet, the five bars should sum to 100%.)

Facet by `Age Group`. The cancer-type strings are very long — make a
short label for each (e.g. `"Lung"`, `"Lymphoid"`, `"Pancreas"`, …) so
the x-axis is readable. Briefly note one age group whose pattern looks
unlike the others, and which cancer type drives the difference.

```{r}

```

## Question 2: Joining `Lahman` tables

For this question we will use the `Lahman` package, which mirrors Sean
Lahman's baseball database. The tables you'll need are:

- `People` — one row per player (`playerID`, `nameFirst`, `nameLast`, …)
- `Batting` — one row per player-year-stint of batting statistics
- `Teams` — one row per team-year (`yearID`, `lgID`, `teamID`, team stats)
- `HallOfFame` — voting results, one row per (player, year, ballot type);
  `inducted == "Y"` means the player was elected
- `AwardsPlayers` — one row per player-award-year

(Quick sanity check: the *primary key* for `Batting` is
`(playerID, yearID, stint)`, since a player can switch teams mid-season
and pick up a new `stint`.)

### Part a — Mutating join with `left_join`

Compute career home runs for every player by grouping `Batting` on
`playerID` and summing `HR`. Sort descending and keep the top 10. Then
attach the player's first and last name from `People` using `left_join`,
and print the result so the names are visible.

```{r}

```

### Part b — Joining on a compound key

Each row of `Batting` belongs to a (player, year, team, league) row in
`Teams`. The shared key here is the *compound key*
`(yearID, lgID, teamID)`.

Use `inner_join` to attach each batter's *team* HR total (call it
`team_HR`) to their `Batting` row on that compound key. Compute each
player-year's share of their team's home runs (`HR / team_HR`).

To keep the question modern (early baseball had tiny HR totals where one
player hitting one homer was the whole team's output), restrict to
seasons where `team_HR >= 100`. Among those, find the 5 player-years with
the *highest* personal share of the team's home runs, and attach the
player's name via `People`.

```{r}

```

### Part c — Filtering join with `semi_join`

Let `hof` be the distinct `playerID`s in `HallOfFame` where
`inducted == "Y"`. Use `semi_join` to keep only the rows of `Batting`
that belong to a Hall-of-Fame player. Report:

- the number of `Batting` rows that survived,
- the number of distinct Hall-of-Fame players who actually have a
  batting record.

```{r}

```

### Part d — Filtering join with `anti_join`

Not every Hall-of-Fame inductee was a batter — the HoF also honors
managers, umpires, and executives. Use `anti_join` to find the
inductees in `hof` who have *no* row in `Batting`. How many are there?

Attach `nameFirst`, `nameLast` from `People` and `category` from
`HallOfFame` to the result, then summarize: how many HoF inductees with
no batting record fall into each `category`?

```{r}

```

## Question 3: Set operations

### Part a — On vectors

From `AwardsPlayers`, build two vectors of distinct `playerID`s:

- `mvp_ids` — players who have won `"Most Valuable Player"`,
- `cy_ids`  — players who have won `"Cy Young Award"`.

Use base R set operations to answer:

- How many players have won *both* awards (`intersect`)? Look up their
  full names in `People` and print the list — you should recognize some.
- How many distinct players have won *either* award (`union`)?
- How many players have won MVP but *not* a Cy Young (`setdiff`)?
- Is `"ruthba01"` (Babe Ruth) in `mvp_ids`? Use `%in%`.

```{r}

```

### Part b — On tables

Set operations on tables compare rows: two rows are "the same" only if
*all* their column values match.

Build two small tables of `(playerID, yearID)` pairs:

- `as_pairs`  — every distinct (`playerID`, `yearID`) appearing in
  `AllstarFull`,
- `mvp_pairs` — every (`playerID`, `yearID`) where the player won
  `"Most Valuable Player"` that year.

Use `intersect`, `union`, and `setdiff` on these tables to answer:

- How many MVP seasons were *also* All-Star seasons for the same player
  (`intersect`)?
- How many distinct (player, year) pairs appear in *either* table
  (`union`)?
- How many MVPs were *not* All-Stars in the year they won
  (`setdiff(mvp_pairs, as_pairs)`)? What might explain those cases?

```{r}

```

## Reflection (10 points)

This section is graded on the substance of your answers, not on getting the
"right" answer. Aim for a few honest sentences each. One-word answers will
not earn full credit.

### 1. Hardest moment (3 points)

Which part of this homework took the longest, or where did you get most
stuck? Describe specifically what was confusing and how you got unstuck.

### 2. Wide vs. tidy (3 points)

In Question 1 you took a wide table with one column per cancer type and
pivoted it to a tidy table with one row per (period, demographic, cancer
type). Name one task that was easier *after* pivoting, and one situation
where you might prefer to keep the original wide layout.

### 3. Choosing a join (2 points)

In Question 2 you used a `left_join`, an `inner_join`, a `semi_join`, and
an `anti_join`. In a sentence or two, describe how you decide which one
to reach for. What question are you really asking when you pick
`semi_join` over `inner_join`?

### 4. A question for class (2 points)

One question you would want to ask in class about pivoting, joins, or set
operations.
