---
title: "DATASCI 306, Spring 2026, Homework 6"
author: "Your name, Your uniqname"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(bench)
library(parallel)
library(foreach)
library(doParallel)
# profvis is used in the Console only (its output is interactive and does not
# knit into a PDF), so we do not load it here.
```

This homework covers material from Lectures 20, 21, and 22: **debugging**
and the three signal levels (`message`, `warning`, `stop`, plus
`traceback` and `browser`); measuring performance
by **profiling** (`profvis`) and **benchmarking** (`bench::mark`); and
**parallel programming** with the `parallel` and `foreach`/`doParallel`
packages (`makeCluster`, `parLapply`, `clusterSplit`,
`foreach` with `%do%`/`%dopar%`), reasoned about with Amdahl's Law.

**Grading.** Same scheme as previous homeworks:

- **Code questions (Q1--Q3): completion.** Attempt every part in good
  faith. Code that runs and shows you engaged with the question earns
  full credit.
- **Reflection: points.** The reflection at the end is graded on the
  substance of your answers (10 points total).

A few parts ask you to run something **in the Console** (profiling output
and some parallel code are awkward or impossible to knit) and then
**write up** what you saw. For those parts, the credit is in your written
description, not in evaluated chunk output.

### Submission checklist

Before submitting, confirm:

- [ ] You replaced the name and uniqname in the YAML header.
- [ ] The document knits to PDF without errors.
- [ ] Every R chunk has been run and shows output where output is expected.
- [ ] You answered the reflection at the bottom of the document.

## Question 1: Debugging and signaling

### Part a --- Finding the bug in a call stack

The file `score_report.R` (in this folder) defines a small four-step
pipeline that should turn a vector of exam scores into a table of
letter-grade counts: `score_report` calls `grade_all`, which calls
`grade_one`, which calls `letter_grade`. 

As written it throws an error. The chunk below loads the file and runs it;
it has `eval = FALSE` so this document still knits:

```{r, eval = FALSE}
source("score_report.R")
score_report(c(95, 83, 72, 60, 88, 91))
# Error in if (score >= 90) { : the condition has length > 1
```

Run the call **in the Console** and use R's debugging tools (start with
`traceback()`) to work out where the error really comes from --- keeping in
mind the Lecture 20 warning that the line an error is *raised* on is not
always where the bug actually *is*. Write a few sentences explaining what
is going wrong, then fix `score_report.R` and show your corrected pipeline
works by setting `eval = TRUE` in the chunk above.

### Part b --- Setting a breakpoint with `browser()`

This bug is sneakier than part a: nothing errors and nothing warns, the
answer is just **wrong**. The function below is meant to return the
*running maximum* of a vector --- at each position, the largest value seen
so far --- so `running_max(c(3, 1, 4, 1, 5))` should give
`c(3, 3, 4, 4, 5)`.

```{r}
running_max <- function(x) {
  best <- 0
  out <- numeric(length(x))
  for (i in seq_along(x)) {
    if (x[i] > best) {
      best <- x[i]
    }
    out[i] <- best
  }
  out
}

running_max(c(-5, -3, -8, -1))   # should be c(-5, -3, -3, -1)
```

The output above is clearly wrong, but there is no error message to read
and no stack to trace --- you need to watch the function's *state* as it
runs. That is what a **breakpoint** is for. In Lecture 20 we used
`browser()`: drop a call to `browser()` inside the function (e.g. as the
first line of the loop body) and R will pause there each time it is
reached, dropping you into an interactive prompt where you can inspect
variables and step with `n` (next), `c` (continue), and `Q` (quit).

**In the Console**, add a `browser()` to `running_max`, run
`running_max(c(-5, -3, -8, -1))`, and step through the first couple of
iterations. Inspect `best` and `x[i]` at the breakpoint. Then write up:
what is the value of `best`, why does the `if` condition never become
`TRUE` for this input, and what one-line change to the function fixes it?

```
Write your findings here (a few sentences).



```

Finally, write the corrected `running_max` in the chunk below and verify
it returns `c(-5, -3, -3, -1)` on the negative input *and* `c(3, 3, 4, 4, 5)`
on `c(3, 1, 4, 1, 5)`.

```{r}

```

### Part c --- Defensive programming with all three signals

Write a function `clean_heights(x)` that cleans a vector of human heights
recorded **in centimeters**, using each of R's three signal levels:

- emit a **`message`** reporting how many values it is cleaning,
- if any value is **negative**, raise a **`warning`** and replace those
  values with `NA`,
- if any value (ignoring `NA`s) is greater than `300`, **`stop`** with an
  informative error (no human is 3 m tall, so this indicates bad data).

Return the cleaned vector. Demonstrate it on `c(170, 165, -5, 182)`
(which should warn and return an `NA`), and show in an `eval = FALSE`
chunk that `clean_heights(c(170, 999))` stops.

```{r}

```

## Question 2: Profiling and benchmarking

The pipeline below computes a few summaries of the spread of a numeric
vector: it centers the values, squares the centered values with the helper
`square_all`, and then reports the count, the variance, and the largest
squared deviation.

```{r}
square_all <- function(x) {
  out <- c()
  for (v in x) {
    out <- c(out, v^2)
  }
  out
}

analyze_spread <- function(x) {
  centered <- x - mean(x)
  sq <- square_all(centered)
  c(n        = length(x),
    variance = sum(sq) / (length(x) - 1),
    max_sq   = max(sq))
}

analyze_spread(rnorm(2000))
```

### Part a --- Profile the pipeline and find the hot spot

`analyze_spread` does several things; only one of them is slow. **In the
Console** (profiling output is interactive and does not knit), profile a
larger run and find out which step dominates:

```{r, eval = FALSE}
library(profvis)
profvis(analyze_spread(rnorm(50000)))
```

Write up which function --- and which *line* of it --- accounts for almost
all of the time, what role `<GC>` plays, and what specifically about that
line makes it slow.

```
Write your findings here.



```

### Part b --- Write a faster replacement

Rewrite the slow helper as `square_all_fast(x)` so the pipeline no longer
has that bottleneck. Verify your version returns the same result as the
original `square_all` on a small input.

```{r}

```

### Part c --- Benchmark

Use `bench::mark` to compare the original `square_all` against your
`square_all_fast` on a sizable input (say `rnorm(20000)`). Report the
result, and using the **median** column (Lecture 20 notes `min`/`median`
are the most trustworthy), state roughly how many times faster your version
is.

```{r}

```

## Question 3: Parallel programming

### Part a --- Amdahl's Law

Amdahl's Law gives the speedup $S$ from parallelizing a fraction $p$ of a
program across $k$ processors:

$$S = \frac{1}{1 - p + p/k}.$$

Write a function `amdahl(p, k)` that returns $S$, then use it to fill in
the speedups for these three scenarios:

| scenario | $p$  | $k$ |
|----------|------|-----|
| A        | 0.90 | 4   |
| B        | 0.90 | 8   |
| C        | 0.50 | 8   |

Comparing A vs B (more processors) and B vs C (more of the program made
parallel), which lever buys more speedup here? Answer in a sentence.

```{r}

```

### Part b --- When does parallelism actually pay off?

Part a's formula assumes parallelizing is free. It is not: starting workers
and shipping each task to them costs time, and that fixed **overhead** only
pays off when each task does enough real work. We will measure where that
break-even point is.

The function below approximates $\pi$ with the Leibniz series
$\pi = 4\sum_{k=0}^{N-1} \frac{(-1)^k}{2k+1}$. The number of terms `n_terms`
is a dial for how much work one call does: more terms means a more accurate
answer and more computation.

```{r}
pi_series <- function(n_terms) {
  k <- 0:(n_terms - 1)
  4 * sum((-1)^k / (2 * k + 1))
}

pi_series(1e6)
```

Computing several independent approximations is embarrassingly parallel.
Set up a 2-worker cluster and **export** `pi_series` to it:

```{r}
cl <- makeCluster(2)
clusterExport(cl, "pi_series")
```

For each value of `n_terms` in `c(10, 1000, 1e5, 1e6, 1e7)`, use
`bench::mark` (with `check = FALSE, memory = FALSE, filter_gc = FALSE`) to
time computing
**eight** approximations *serially* with `lapply(1:8, ...)` against *in
parallel* with `parLapply(cl, 1:8, ...)`. Collect the **median** of each
into a table with columns `n_terms`, `serial`, `parallel`, and
`speedup = serial / parallel` (a `map`/`map_dfr` over the `n_terms` values
is a clean way to build it). Then `stopCluster(cl)`.

Report your table. In a few sentences: at small `n_terms`, is the parallel
version faster or *slower* than serial, and why? As `n_terms` grows, what
does the speedup approach, and why can't it exceed that? Connect both ends
to the overhead term that Amdahl's Law from part a leaves out.

```{r pi-sweep, cache = TRUE}

```

### Part c --- Split--apply--combine with `foreach`

Not every problem can be easily parallelized using tools like `parLapply`. To compute a single
aggregate across a vector in parallel, the pattern is **split** the data
into chunks, **apply** a per-chunk computation, then **combine** the
chunk results.

In part b each task computed a *whole* approximation of $\pi$. Here we go
the other way: take a **single** Leibniz approximation and split *it* across
workers. Because the approximation is a sum over the term indices
$k = 0, 1, \ldots, N-1$, and a sum can be broken into pieces, we can sum
each block of indices on its own worker and then add the block totals.

Implement this split--apply--combine for `N <- 1e7`:

- **Apply:** write `leibniz_chunk(ks)` returning the partial sum
  $\sum_{k \in ks} \frac{(-1)^k}{2k+1}$ over a vector of term indices `ks`.
- **Split:** make a 2-worker cluster and use `clusterSplit(cl, 0:(N - 1))`
  to break the indices into one chunk per worker.
- **Combine:** iterate over the chunks with `foreach`, applying
  `leibniz_chunk` to each and passing the `+` operator as `.combine` so the
  partial sums are added; multiply the combined total by 4 to get $\pi$.

Use `%do%` so the document knits reliably (register the cluster with
`registerDoParallel(cl)` and switch to `%dopar%` if you want the chunks to
actually run on the workers). Verify your result matches `pi_series(N)`
from part b, then `stopCluster(cl)`.

```{r}

```

## Reflection (10 points)

This section is graded on the substance of your answers, not on getting
the "right" answer. Aim for a few honest sentences each. One-word answers
will not earn full credit.

### 1. Hardest moment (3 points)

Which part of this homework took the longest, or where did you get most
stuck? Describe specifically what was confusing and how you got unstuck.

### 2. A bug of your own (3 points)

Think of a bug you have hit in your own R code this term (a confusing
error, a wrong answer, a warning you ignored). Which of the Lecture 20
strategies --- searching, shrinking to a small example, locating it in the
call stack with `traceback()`, or pausing to inspect state with a
`browser()` breakpoint --- would have helped you find it fastest, and why?

### 3. To parallelize or not (2 points)

In Question 3 you saw both a real speedup and a case where parallelism made
things slower. In your own words, what has to be true about a task before
reaching for parallel code is worth it?

### 4. A question for class (2 points)

One question you would want to ask in class about debugging, profiling, or
parallel programming.
